---
title: "RPhenoscape Tutorial"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
## 1. Introduction

The desired input of many comparative studies is a matrix that includes all the data available for a set of traits and a focal taxon.  Bringing similar types of trait data together across studies is notoriously difficult, but is straightforward with ontology annotations.  Trait data that are tagged using similar ontology terms can be automatically called using R functions. 

This workshop uses trait data for vertebrates that has been annotated (tagged) with ontology terms.  These data are accessed from the Phenoscape Knowledgebase (KB).  In this lesson we use R functions to obtain and view character matrices from the KB, obtain synthetic matrices of inferred presence/absence characters, and understand the meaning and relations of ontology terms. The lesson assumes basic knowledge of R and RStudio.

## 2. Browse and filter data in the KB to understand the scope of available data.

You would like to understand the evolution of morphological traits in a group, and you need a matrix to map to your (typically molecular) phylogenetic tree.  You come to the Phenoscape KB to see what types of data are available across the members of your clade.  In this exercise, you are an ichthyologist, interested in catfishes, specifically the bullhead catfishes (Ictaluridae).
You begin by querying ‘Siluriformes’  in the Phenoscape KB faceted browsing page http://beta.phenoscape.org/#/facet. You immediately see the scope of the data:

From left to right across the top tabs, for ‘Siluriformes’:
- There are 4,537 unique phenotypes
- There are 292,356 taxon annotations 
- There are 1,452 siluriform taxa with data in the KB 
- There are 25 publications in the KB that include catfish phenotypes
- There are 17,883 candidate genes linked to the phenotypes

Taxon annotations are the ontology terms tagged to free text descriptions of phenotypes for each siluriform taxon.  Click on the ‘Sources’ box on the right hand side of each row to see the publication source and free text description of the phenotype from which the ontology annotation was derived.  

You are specifically interested in Ictaluridae (bullhead catfishes), and you click on that family in the faceted browsing interface.  Now filter using ‘fin’ as the anatomical entity and click ‘parts’.  

Download these taxa and their morphological data using the link provided.  

Go to the publications tab, which shows nine publications that contain fin data.  Note that any of these studies, e.g. Lundberg (1992), can be entered under ‘publications’ in faceted browsing, but the original matrix cannot be downloaded. 

## 3. Retrieve term info using RPhenoscape

Now we will repeat some of the above steps using RPhenoscape, beginning with a query for Ictaluridae (bullhead catfishes) and ‘fin’ (including parts) to retrieve the list of studies that contain fin characters for this taxon.  

First let’s look up information about the query terms. You can use the term detail functions ```pk_anatomical_detail()```, ```pk_taxon_detail()```, ```pk_phenotype_detail()```, and ```pk_gene_detail()``` to retrieve an ontology term’s ID, label, and definition. 

For example, view the details for the anatomy term “fin”:

```{r}
```pk_anatomical_detail("fin")```
```
For a specific taxon, such as Siluriformes, you can view the taxon rank, term label, whether or not the taxon is extinct, its common name, and ID: 

```{r}
```pk_taxon_detail("Siluriformes")```

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
